A general and efficient representation of Ancestral Recombination Graphs

### 2022-04-06 paper.tex
Experiment with using named labels for section refs

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{calc,positioning}

@@

\linenumbers
\title{What is an Ancestral Recombination Graph?}
% First authors
\author{Author McAuthorface}
% Corresponding

@@

Recent breakthroughs in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified}
and data representation~\citep{kelleher2016efficient}
have raised the genuine prospect of genealogical analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these inferred ancestries as input have
begun to appear~\citep{osmond2021estimating} and many more are sure to
follow~\citep{harris2019database}. This vibrant research area, however,
has a significant difficulty, which, if not addressed, will substantially hamper
progress. We currently lack a well-defined data-model and shared terminology
to discuss recombining genealogies,
leading to basic errors in statements about what different inference methods
produce as output (see the XXX section below), as well as presenting serious
difficulties in either comparing the outputs of different methods or
(from a users perspective) using multiple inference methods in an analysis.

@@

Fortunately, it is straightforward to generalise the classical Griffiths
EARG encoding to both simplify implementations and to encompass any
possible pattern of genetic inheritance.
The only change that we require is to associate
the details of inheritance intervals with \emph{edges} rather than
as breakpoints associated with recombination events and
edge ordering rules. This is illustrated in Fig~\ref{fig-arg-data-structure}B
where this alternative edge annotation formulation is
shown. By associating the interval $(0, x)$ with edge $(3,4)$
and the interval $(x, L)$ with $(3, 5)$ we make the encoding
unambiguous: we can list the edges in any order, and it
will describe the same set of local genealogical trees.
This approach is also more flexible, because any pattern
of inheritance can be described by these intervals. For
example, if a gene conversion event occurred where a segment
between between positions $x$ and $y$ was inherited by
node $3$ from $5$, we would state this by associating
the intervals $\{(0, x], (y, L]\}$ with $(3,4)$ and
$\{(x, y]\}$ with $(3,5)$.
This change is straightforward, involves no loss of information
and (as we see in later sections)
greatly increases both the expressivity and computational
power of ancestral recombination graph data structures.

@@

To distinguish from the classical event-based encodings,
we call this encoding of an ARG data
structure a ``Genome ARG'', or GARG.
Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a GARG.
Fig.~\ref{fig-arg-in-pedigree}a shows the genomes
embedded in an inbred diploid pedigree, in which individuals occur at
discrete generations and carry two genomes. Each individual (labelled
$i_1$ to $i_8$) therefore has a paternal genome (on the right) and a maternal one
(on the left), although the pedigree is only shown for concreteness, and
there is no inherent limitation on mating system, ploidy, or age structure.
A genome is inherited from one of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $i_1$ has two genomes \textsf{A} and \textsf{B},
inherited from parents $i_3$ and $i_4$. Genome \textsf{A} is the product of
recombining $i_3$'s two genomes \textsf{E} and \textsf{F} at position 2,
whereas \textsf{B} was inherited directly from $i_4$'s genome \textsf{G} without
recombination. Fig.~\ref{fig-arg-in-pedigree}b shows the bare gARG, illustrating
that the patterns of
transmission between genomes can be followed without necessarily knowing
how they are combined within individuals. This ``genome-perspective''
simply depicts the ancestry of the sample genomes $A$--$D$. The graph
topology along with the genome coordinate annotations on the
edges then uniquely defines the trees at every position
along the genome (Fig.~\ref{fig-arg-in-pedigree}c).

Formally, we can define a GARG as a set of edges $E$, where each
element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
also phrase this in terms of the ancestral material of a sample---see
the next section). Deriving the local tree at a point $x$
follows largely the same pattern as for an EARG but is somewhat
simpler because coordinate information is directly associated with
edges. Starting with the sample nodes $S$ we traverse
upwards through the graph from child to parent. At each node $u$, we find an
edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$ (if it
exists) and move to the next node $p$.

The focus on genomes rather than events has some interesting
and important ramifications. Firstly, a gARG may contain nodes that
do not correspond to any particular event in the ancestry of the samples.
For example, node \textsf{H} is the direct ancestor of \textsf{D} in
Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
coalescence or a recombination. Such nodes would usually be removed
from the graph so that \textsf{D} descends directly from \textsf{K} (see
section XXX) but there is no necessity for this from a representational
perspective. This ability can be useful in practice and could be required if,
% Do we need to spell this out?
for example, we have a deeply sequenced pedigree of
haplodiploids. % e.g. Honeybees? Be great to cite an example!

Secondly, unlike in an eARG, nodes in a gARG are not classified into different
types. Indeed a single node can have both multiple children
and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
for instance, node \textsf{A} were to have two children.
% It is a shame that this is not actually shown in fig 3, but I don't think
% we can do this unless we have 5 generations
Moreover, a single node can have more than 2 children, and (as we shall see in
section XXX), more than 2 parents. Another way to state this is that transitions
between nodes in a gARG can encapsulate multiple events in an eARG.

@@

\section*{Ancestry resolution}
% This is a rough first pass. Will clarify terminology later.
The definition of an ARG as a data structure in terms of
individuals' monoploid genomes
and their direct patterns of inheritance allows us to describe arbitrarily
complex relatedness structures, free from the limitations imposed by
coalescent approximations.
An ARG is an intrinsically retrospective
structure, defined with respect to the ancestry of
a set of sampled genomes.
The graph structure (the nodes and their connectivity through edges)
captures a great deal of information about this ancestry,
but on its own it cannot tell us about the precise relationships between
the samples along the genome: do to this, we must construct the
local (or ``marginal'') trees.

@@

\section*{ARG simplification}

@@

The ideas of ``meaningful'' local tree topologies revolve around the
presence of ``unary'' nodes: a node in the tree that has exactly
one child. Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a Wright-Fisher
model, shown both as a graph on the left, and the equivalent set of 6
local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
\textsf{A} in the penultimate tree, highlighted in yellow and covering genomic positions
60 to 70. On this path, before $A$ meets its common ancestor with the other samples at
node \textsf{Q}, it passes through  \textsf{E},  \textsf{J},  \textsf{M},  \textsf{N} and 
\textsf{P}, which in this region of the genome are all unary nodes. This reflects the
passage of this span of genome through the ARG towards the root (see section XXX).
Such unary nodes in a local tree are not usually considered important,
because they provide very little information and are not
detectable by information from the tree. % weak
The nodes in a tree usually mark the points where samples find a
 \emph{common} ancestor, such as node \textsf{Q} in the highlighted
tree. For many purposes, we would therefore consider a path directly
joining sample \textsf{A} to node \textsf{Q} to be of equivalent status to the path
containing the intermediate unary nodes. Simplification is the process of removing
these less informative and (for most computational purposes) inefficient nodes
from the local trees and the corresponding annotated graph.

@@

\section*{ARG inference methods}
Using the terminology developed here to classify ARGs, we now review methods developed
to explicitly \emph{infer} ARGs from sequencing data.

>>>>

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{calc,positioning}
\usepackage{hyperref}

@@

\linenumbers
\title{A general and efficient representation of Ancestral Recombination Graphs}
% First authors
\author{Author McAuthorface}
% Corresponding

@@

Recent breakthroughs in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified}
and data representation~\citep{kelleher2016efficient}
have raised the genuine prospect of genealogical analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these inferred ancestries as input have
begun to appear~\citep{osmond2021estimating} and many more are sure to
follow~\citep{harris2019database}. This vibrant research area, however,
has a significant difficulty, which, if not addressed, will substantially hamper
progress. We currently lack a well-defined data-model and shared terminology
to discuss recombining genealogies, leading to basic errors in statements
about what different inference methods produce as output (see the
``\nameref{ARG_inference_methods}'' section), as well as presenting serious
difficulties in either comparing the outputs of different methods or
(from a users perspective) using multiple inference methods in an analysis.

@@

Fortunately, it is straightforward to generalise the classical Griffiths
eARG encoding to both simplify implementations and to encompass any
possible pattern of genetic inheritance.
The only change that we require is to associate
the details of inheritance intervals with \emph{edges} rather than
as breakpoints associated with recombination events and
edge ordering rules. This is illustrated in Fig~\ref{fig-arg-data-structure}B
where this alternative edge annotation formulation is
shown. By associating the interval $(0, x)$ with edge $(3,4)$
and the interval $(x, L)$ with $(3, 5)$ we make the encoding
unambiguous: we can list the edges in any order, and it
will describe the same set of local genealogical trees.
This approach is also more flexible, because any pattern
of inheritance can be described by these intervals. For
example, if a gene conversion event occurred where a segment
between between positions $x$ and $y$ was inherited by
node $3$ from $5$, we would state this by associating
the intervals $\{(0, x], (y, L]\}$ with $(3,4)$ and
$\{(x, y]\}$ with $(3,5)$.
This change is straightforward, involves no loss of information
and (as we see in later sections)
greatly increases both the expressivity and computational
power of ancestral recombination graph data structures.

@@

To distinguish from the classical event-based encodings,
we call this encoding of an ARG data
structure a ``Genome ARG'', or gARG.
Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG.
Fig.~\ref{fig-arg-in-pedigree}a shows the genomes
embedded in an inbred diploid pedigree, in which individuals occur at
discrete generations and carry two genomes. Each individual (labelled
$i_1$ to $i_8$) therefore has a paternal genome (on the right) and a maternal one
(on the left), although the pedigree is only shown for concreteness, and
there is no inherent limitation on mating system, ploidy, or age structure.
A genome is inherited from one of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $i_1$ has two genomes \textsf{A} and \textsf{B},
inherited from parents $i_3$ and $i_4$. Genome \textsf{A} is the product of
recombining $i_3$'s two genomes \textsf{E} and \textsf{F} at position 2,
whereas \textsf{B} was inherited directly from $i_4$'s genome \textsf{G} without
recombination. Fig.~\ref{fig-arg-in-pedigree}b shows the bare gARG, illustrating
that the patterns of
transmission between genomes can be followed without necessarily knowing
how they are combined within individuals. This ``genome-perspective''
simply depicts the ancestry of the sample genomes $A$--$D$. The graph
topology along with the genome coordinate annotations on the
edges then uniquely defines the trees at every position
along the genome (Fig.~\ref{fig-arg-in-pedigree}c).

Formally, we can define a gARG as a set of edges $E$, where each
element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
also phrase this in terms of the ancestral material of a sample---see
the next section). Deriving the local tree at a point $x$
follows largely the same pattern as for an eARG but is somewhat
simpler because coordinate information is directly associated with
edges. Starting with the sample nodes $S$ we traverse
upwards through the graph from child to parent. At each node $u$, we find an
edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$ (if it
exists) and move to the next node $p$.

The focus on genomes rather than events has some interesting
and important ramifications. Firstly, a gARG may contain nodes that
do not correspond to any particular event in the ancestry of the samples.
For example, node \textsf{H} is the direct ancestor of \textsf{D} in
Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
coalescence or a recombination. Such nodes would usually be removed
from the graph so that \textsf{D} descends directly from \textsf{K} (see
the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
a representational perspective. This ability can be useful in practice and could be
% Do we need to spell this out?
required if, for example, we have a deeply sequenced pedigree of haplodiploids.
% e.g. Honeybees? Be great to cite an example!
% Note by Yan - if we actually have the sequence, then this turns into
% a sample node, so would be normally included anyway. Perhaps, therefore
% we should talk about these "pass though" nodes in the context
% of internal samples.

Secondly, unlike in an eARG, nodes in a gARG are not classified into different
types. Indeed a single node can have both multiple children
and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
for instance, node \textsf{A} were to have two children.
% It is a shame that this is not actually shown in fig 3, but I don't think
% we can do this unless we have 5 generations
Moreover, a single node can have more than 2 children, and (as we shall see in
the ``\nameref{ARG_simplification}'' section, more than 2 parents. Another way to
state this is that transitions between nodes in a gARG can encapsulate multiple
events in an eARG.

@@

\section*{Ancestry resolution}\label{Ancestry_resolution}
% This is a rough first pass. Will clarify terminology later.
The definition of an ARG as a data structure in terms of
individuals' monoploid genomes
and their direct patterns of inheritance allows us to describe arbitrarily
complex relatedness structures, free from the limitations imposed by
coalescent approximations.
An ARG is an intrinsically retrospective
structure, defined with respect to the ancestry of
a set of sampled genomes.
The graph structure (the nodes and their connectivity through edges)
captures a great deal of information about this ancestry,
but on its own it cannot tell us about the precise relationships between
the samples along the genome: do to this, we must construct the
local (or ``marginal'') trees.

@@

\section*{ARG simplification}\label{ARG_simplification}

@@

The ideas of ``meaningful'' local tree topologies revolve around the
presence of ``unary'' nodes: a node in the tree that has exactly
one child. Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a Wright-Fisher
model, shown both as a graph on the left, and the equivalent set of 6
local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
\textsf{A} in the penultimate tree, highlighted in yellow and covering genomic positions
60 to 70. On this path, before $A$ meets its common ancestor with the other samples at
node \textsf{Q}, it passes through  \textsf{E},  \textsf{J},  \textsf{M},  \textsf{N} and 
\textsf{P}, which in this region of the genome are all unary nodes. This reflects the
passage of this span of genome through the ARG towards the root (see the
``\nameref{Ancestry_resolution}'' section).
Such unary nodes in a local tree are not usually considered important,
because they provide very little information and are not
detectable by information from the tree. % weak
The nodes in a tree usually mark the points where samples find a
 \emph{common} ancestor, such as node \textsf{Q} in the highlighted
tree. For many purposes, we would therefore consider a path directly
joining sample \textsf{A} to node \textsf{Q} to be of equivalent status to the path
containing the intermediate unary nodes. Simplification is the process of removing
these less informative and (for most computational purposes) inefficient nodes
from the local trees and the corresponding annotated graph.

@@

\section*{ARG inference methods}\label{ARG_inference_methods}
Using the terminology developed here to classify ARGs, we now review methods developed
to explicitly \emph{infer} ARGs from sequencing data.
